---
title: "Template for Analysis & APA Style"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this section, I want to take everything we have highlighted in the previous two walkthroughs and lay out a template for ANOVA. Think of this as ANOVA in 9 easy steps. FWIW I also have a video link posted on CANVAS that invites you to do a problem with me in real time that reiterates much of what's covered here. This walkthrough in many ways mirrors that video and is nearly identical to the **qmd** that accompanies the video (all of which is to say if you're a more visual person, the video may be the way to go).

## Step 0: Loading in the packages

```{r}
pacman::p_load(psych, tidyverse, cowplot, emmeans, ez)
```

## Step 1: Data wrangling

> -   import your data
>
> -   is the data in the right format?
>
> -   do you need to make any adjustments to the data frame?

```{r step 0 code}
library(readr)
dataset <- read_csv("http://tehrandav.is/courses/statistics/practice_datasets/factorial_ANOVA_dataset_no_interactions.csv")

# unnecessary here, but if performing planned contrasts you need to factorize predictors (IVs)
dataset$Presentation <- as.factor(dataset$Presentation)
dataset$Lecture <- as.factor(dataset$Lecture)

dataset
```

-   In this case the data looks to be in the correct format (long). Since I am not doing planned contrasts I don't need to worry about *factorizing* my IV columns.

## Step 2: Look at your data

> -   plot the data: I like to do histograms by condition as well as a means plot (interaction plot) and a participant number grid to get a feel for what my distribution of scores looks like. (Note I'm not using this to make decisions on my data model assumptions!)

```{r step 1 code histogram plots}
ggplot(data = dataset, aes(x = Score)) +
    geom_histogram(fill = "gray", color = "black") +
    facet_grid(rows = vars(Lecture), cols = vars(Presentation))
```

I also take a quick glance at the means of my groups (interaction plot)

```{r step 1 code means / boxplots}
ggplot(data = dataset, aes(x = Lecture, y = Score, group = Presentation, color = Presentation)) +
    stat_summary(fun.data = "mean_se",
                 geom = "pointrange") +
    stat_summary(fun = "mean",
                 geom = "line",
                 lty="solid")
```

And finally I take a look at the number of participants in each group (cell) to ensure that my design is roughly balanced.

```{r eztable number of participants}
ez::ezDesign(dataset, x = Lecture, y = Presentation)
```

## Step 3: Build your model

> -   build the ANOVA model
>
> -   note that this week we are only focusing on the additive model.

```{r step 3 code to specify model}

# additive model (this week)
aov_model <- lm(Score~Lecture + Presentation, data = dataset)
```

## Step 4: Check your model assumptions

> -   check the model assumptions (normality and homogeneity)

```{r step 4 checkimng your model assumptions}

# visual:
aov_model %>% performance::check_model(panel = FALSE)

# test of normality (this is the Shapiro Test):
aov_model %>% performance::check_normality()

# test of homogeneity
aov_model %>% performance::check_homogeneity()



```

[***Note***]{.underline}*: that if you wanted to check normality using the bootstrapping method from the video, there is a mistake (I forgot an important step). FWIW, this would be better than Shapiro-test but it's much more involved*

```{r}
# note that if you wanted to check normality using the bootstrapping method from the video, these is a mistake (I forgot an important step).

# 1. build a distribution of skews (note that because this is random, your values won't equal mine):
distribution_skews <- DescTools::Skew(x = aov_model$residuals,
                ci.type = "bca", 
                method = 2, 
                conf.level = 0.95, 
                R = 1000)
# 2. calculate the standard error:
standard_error_of_skews <- (distribution_skews[3]-distribution_skews[2])/3.92

# 3. take obtained skew and divide by standard_error_of_skews
psych::skew(aov_model$residuals) / standard_error_of_skews %>% 
  unname()

# 4. compare this value against critical values from Kim (2013) method, repeat for kurtosis

# ----- OR ------- use my custom function that handles all of the above

source("http://tehrandav.is/courses/statistics/custom_functions/skew_kurtosis_z.R")

aov_model$residuals %>% skew_kurtosis_z()
```

### SUMMARY:

-   Test of normality and homogeneity don't suggest troublesome any violations

## Step 5: Run the analysis (ANOVA)

```{r step 5 code}
pacman::p_load(sjstats)
sjstats::anova_stats(aov_model)[1:8]
```

### SUMMARY

-   Main effect Lecture: F(2, 30) = 3.83, p = .03, np2 = .20
-   Main effect Presentation: F(1, 30) = 22.96, p \< .001, np2 = .43
-   remember that since this is a factorial ANOVA I report my partial eta square for effect size.

## Step 6: Run additional tests based what your ANOVA outcomes

> -   deal with the ANOVA outcomes ONE AT A TIME!!!
>
>     -   simple effects ANOVA (if there is an interaction... more on this next week)
>
>     -   post hoc analyses
>
>     -   or if you have planned contrasts run those here
>
> -   make a quick summary of each result
>
>     -   include any relevant means or tests (t-tests, p vals, etc)

Here we are performing posthocs. We'll have two sets (although for now the post hoc test on Presentation is redundant.

### Post hoc analysis for Lecture

```{r step 5 code - main effect 1 post hocs}
emmeans(aov_model, spec = ~Lecture) %>% pairs(adjust = "tukey")
```

#### SUMMARY

-   tukeys suggests difference between Phys and Social t(30) = 2.71, p = .029
-   Phys \> Soc

### Post hoc analysis for Presentation

```{r step 6 code - main effect 2 post hocs}
emmeans(aov_model, spec = ~Presentation) %>% pairs(adjust = "tukey")
```

#### SUMMARY

-   tukeys suggests difference between Comp and Stand
-   Comp \> Stand
-   t(30) = 4.792, p \< .001

## Step 7: Create a camera ready plot

I really go into detail with this in the 3rd video (see CANVAS). For now, a few considerations:

> -   publication plots by default need to be in black/white/grayscale
>
> -   the legend needs to fit within the axes of the plot
>
> -   when plotting multiple factors on a single plot, its useful to differentiate data series by `shape`, `fill`, `color`, or `linetype` depending on the plot. Sometimes its best to use these aesthetics in combination with one another.
>
> -   Include a figure caption in the text below the plot output.

**For example:**

```{r step 7 code camera ready plot}
# lecture on x-axis, score on y-axis, data groups by presentation:
ggplot(data = dataset, aes(x = Lecture, y = Score, group = Presentation)) +
    # pointrange (mean ± se) with different shapes by presentation type:
    stat_summary(fun.data = "mean_se",
                 geom = "pointrange",
                 aes(shape = Presentation),
                 size = 1) +
  # add lines (means), with different linetypes by presentation
    stat_summary(fun = "mean",
                 geom = "line",
                 aes(linetype = Presentation)) +
  # theme cowplot gets us a quick APA
    theme_cowplot() +
  # custom labels to make my axes convey more information
  # if you con't include this the axis just contain the headers (Lecture, Presentation)
    labs(y="Test Score", x = "Lecture Subject") +
  # adjusting the location of the legend:
    theme(legend.position = c(.05,.9))
```

*Figure 1.* Test scores as a function of subject of lecture and presentation type. Errors bars are standard error

## Step 8: write ups

> -   you can perform a write up in Word (or you word processor of choice)
>
> -   or you can do a write-up in Rstudio
>
>     -   you can use Markdown and MathJax (LaTeX) to do this.

Before starting, let's get our summary statistics (mean and standard errors) to include in the write up.

**Lecture:**

```{r}
Rmisc::summarySE(dataset, measurevar = "Score", groupvars = "Lecture")
```

**Presentation**

```{r}
Rmisc::summarySE(dataset, measurevar = "Score", groupvars = "Presentation")
```

Taking these summary statistics in addition to the ANOVA table summary I outlined above, I can perform a detailed write up. Be sure to include

-   all tests run (ANOVA, post hocs, and their appropriate statistics

-   information about group means and standard errors.

-   explicitly refer to the Figure.

-   This should follow a logical flow; in this case after presenting the general ANOVA, focus on post-hocs for one main effect and explain that info completely before moving onto the other.

**For example:**

> We submitted our data to a 2 (Presentation) x 3 (Lecture) Analysis of Variance. Our ANOVA revealed a significant main effect for Presentation, $F$(1, 30) = 22.96, $p$ \< .001, $\eta_p^2$ = .43. Scores for groups given the computer presentation ($M ± SE$: 41.33 ± 1.47) were greater than scores for the standard presentation (29.56 ± 2.22).Our ANOVA revealed a significant main effect for Lecture, $F$(2, 30) = 3.83, $p$ = .03, $\eta_p^2$ = .20. Post hoc Tukey pairwise analysis revealed that the mean score of participants in the Physical (40.00 ± 3.12) group was greater than the score of the Social group (30.83 ± 2.69), $t$(30) = 2.71, $p$ = .029. See Figure 1 for group comparisons.
