---
title: "Factorial RM-ANOVA and simple effects"
---

Let's ramp-up the complexity. Here is a dataset testing Recall over three days as a function of depth of processing (Lo, Med, High). In this case we have two within factors.

## loading in the necessary packages and data:

```{r}
pacman::p_load(tidyverse,
               afex,
               cowplot,
               emmeans)

source("http://tehrandav.is/psyc7014/custom_functions/withinSummary.R")

factorial_df <- read_csv("http://tehrandav.is/psyc7014/practice_datasets/withinEx3.csv")

factorial_df
```

One thing to note is that **Subject** is being treated as a \<dbl\> number. It should be treated as a factor.

```{r}
factorial_df$Subject <-  as.factor(factorial_df$Subject)
```

## plotting the data

### individual subjects plots: Spaghetti

Since I have two factors here, one strategy may be to **facet** my spaghetti plots. In this case I am creating seperate spaghetti plots for each level of **ProcessDepth:**

```{r}
p <- ggplot(data = factorial_df, aes(x=Day, y=Recalled, group = Subject)) +
  geom_point(aes(col=Subject)) + 
  geom_line(aes(col=Subject)) +
  facet_wrap(~ProcessDepth)

plotly::ggplotly(p)
```

FWIW, on thing that this plot has clued my in on is that the order of my **Processing Depth** levels is High, Lo, Med. I might actually want that to be Lo, Med, High:

```{r}
factorial_df$ProcessDepth <- fct_relevel(factorial_df$ProcessDepth, 
                                     c("Lo","Med","High")
                              )   
```

and now re-plotting...

```{r}
p <- ggplot(data = factorial_df, aes(x=Day, y=Recalled, group = Subject)) +
  geom_point(aes(col=Subject)) + 
  geom_line(aes(col=Subject)) +
  facet_wrap(~ProcessDepth)

plotly::ggplotly(p)
```

#### means plots

Again, we need to make the appropriate correction for our error bars. In this case the number of within groups is the number of cells that is formed by crossing Day (1,2,3) and Processing Depth (Lo, Med, Hi). As this is a 3 \* 3 design there are 9 cells, or `nWithinGroups=9`. As before this is handled auto-magically in `summarySEwithin2`, or our custom function `withinSummary`

```{r}
descriptiveStats <- withinSummary(data=factorial_df,
                                  measurevar = "Recalled",
                                  withinvars = c("Day","ProcessDepth"),
                                  idvar = "Subject")


p <- ggplot(descriptiveStats$Corrected, aes(x = Day,
                                            y = Recalled, 
                                            group=ProcessDepth)) +
  geom_point(size=2.5, aes(shape = ProcessDepth)) + 
  geom_line(size=1, aes(linetype=ProcessDepth)) +
  geom_errorbar(aes(ymin=Recalled-se, ymax=Recalled+se), width=0.1) +
  theme_cowplot() +
  theme(
    axis.title = element_text(size = 16, face = "bold", lineheight = .55),
    axis.text = element_text(size = 12),
    legend.title = element_text(size = 12, face = "bold"),
    legend.position = c(.25,.75)) +
  scale_color_manual(values=c("black","grey50")) + 
  xlab("Lecture") + 
  ylab ("Score") +
  theme(plot.margin=unit(c(.25,.25,.25,.25),"in")) + 
  
  # stack legend boxes horizontally:
  theme(legend.box = "horizontal")

show(p)
```

### running the omnibus ANOVA:

```{r}
omnibus_aov <- afex::aov_ez(id = "Subject", 
                            dv = "Recalled", 
                            data = factorial_df,
                            between = NULL,
                            within = c("ProcessDepth","Day"),
                            type = 3, 
                            return = "afex_aov", 
                            include_aov = TRUE, 
                            anova_table=list(es = "pes"))

summary(omnibus_aov)
```

Here we have two main effects and an interaction. Let's unpack the interaction by taking a look at whether `Recall` increases over successive days.

### running the simple effects ANOVAs

#### Recommended method (assumed violation of spherecity)

The Cohen chapter assigned this week recommends that **one avoid the use of pooled error terms when performing simple effects analysis of within-subjects ANOVA**. This is definitely the case when the spherecity assumption is violated, but as a general rule, even when the assumption is not technically violated, deviations from spherecity can artificially inflate Type I error. So in this case, simple follow up ANOVAs that refer to their own error terms are justified. That said as Type I error goes down, Type II goes up. Something to consider.

One way to do this would be, as we have done in the past, separate out the data and run separate simple effects ANOVAs using `emmeans`. **This time, we DO NOT use the error term from the omnibus ANOVA**, we treat each simple effects ANOVA as self contained. This can be accomplished using `joint_tests()` from `emmeans` telling it how we want to separate our original model. As before, we need to add `model="multivariate"` to our call (this tells `R` to use the simple effects error term and not the omnibus).

Here I'm taking a look at the effect of Day on each level of Processing Depth. A better way of saying this is whether you see different rates of learning over Days depending on what level of Processing is involved.

```{r}
joint_tests(omnibus_aov, by = "ProcessDepth", model="multivariate")
```

#### Assuming spherecity

If we wanted to throw Cohen's caution to the wind, we would simply run the simple effects tests with a `univariate` model. Note that this method attempts to deal with potential violations of our spherecity assumptions by using Satterthwaite degrees of freedom. Keep in mind, we are less protected from Type 1 error in this case.

```{r}
joint_tests(omnibus_aov, by = "ProcessDepth", model="univariate")
```

In both cases we find an effect for `Day` on the `High` and `Med` `ProcessDepth`.

As a follow up we would run the pairwise tests. Here's the multivariate method:

```{r}
# multivariate method
emmeans(omnibus_aov, 
        by = "ProcessDepth", 
        specs = pairwise ~ Day, 
        model="multivariate", 
        adjust="tukey")
```

and the univariate (pooled error term):

```{r}
emmeans(omnibus_aov, 
        by = "ProcessDepth", 
        specs = pairwise ~ Day, 
        model="univariate", 
        adjust="tukey")
```

### Writing this up

... To test this hypothesis we ran a 3 (Process Depth) × 3 (Day) repeated-measures ANOVA. Where there were violations of spherecity, we used Greenhouse-Geisser corrected degrees of freedom. This analysis revealed a significant interaction, $F$(2.86, 25.78) = 10.64, $p$ \< .001, $\eta_p^2$ = .54. To test this interaction we ran seperate simple effects ANOVA for Day on each level of Process Depth. A simple effect for Day was found on High processing depth, $F$(2, 9) = 32.05, $p$ \< .001, where there were statistically significant increases in performance from each Day to the next (Tukey HSD, $p$ \< .05). A simple effect for Day was also found on Med processing depth, $F$(2, 9) = 30.15, $p$ \< .001. While scores increased across the 3 days, only Day 3 was significantly greater than the other two ($p$s \< .05). No simple effect was observed for Day in the Lo processing depth condition.

See the next walkthrough for how to do this in SPSS.
